---
title: "purrr <-> base R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{purrr <-> base R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
options(tibble.print_min = 6, tibble.print_max = 6)
```

# Introduction

This vignette compares purrr functionals to their base R equivalents, focussing primarily on the map family and related functions. This helps those familiar with base R understand better what purrr does, and shows purrr users how you might express the same ideas in base R code. We’ll start with a rough overview of the major differences, give a rough translation guide, and then show a few examples.

```{r setup}
library(purrr)
library(tibble)
```

## Key differences

There are two primary differences between the base apply family and the purrr map family: purrr functions are named more consistently, and more fully explore the space of input and output variants.

* purrr functions consistently use `.` as prefix to avoid function arguments
  mismatches. Base functions use a variety of techniques including upper case
  (e.g. `lapply(X, FUN, ...)`) or require anonymous functions (e.g. `Map()`). 
  See <https://adv-r.hadley.nz/functionals.html#argument-names> for more 
  details.

* All map functions are type stable: you can predict the type of the output
  using little information about the inputs. In contrast, the base functions
  `sapply()` and `mapply()` automatically simplify.

* The map family functions all start with the data, followed by the function, 
  then any additional constant argument. Most apply functions also follow this
  pattern, but `mapply()` starts with the function, and `Map()` has no way
  to supply additional constant arguments.
  
* purrr functions provide all combinations of input and output variants, and
  include variants specifically for the common two argument case. 

* The map family all support an alternative succinct function definition syntax 
  that helps with the most common cases.


## Direct translations

The following sections give a high-level translation between base R commands and their purrr equivalents. See function documentation for the details.

### `Map` functions

Here `X` denotes a vector or a list and `FUN` denotes a function

Output | Input | Base R | purrr
-------|-------|--------|-----------
List | 1 vector | `lapply()` | `map()`
List | 2 vectors | `mapply()`, `Map()` | `map2()`
List | >2 vectors | `mapply()`, `Map()` | `pmap()`
Atomic vector of desired type | 1 vector | `vapply()` | `map_lgl()` (logical), `map_int()` (integer), `map_dbl()` (double), `map_chr()` (character), `map_raw()` (raw)
Atomic vector of desired type | 2 vectors | `mapply()`, `Map()`, then `is.*()` to check type | `map2_lgl()` (logical), `map2_int()` (integer), `map2_dbl()` (double), `map2_chr()` (character), `map2_raw()` (raw)
Atomic vector of desired type | >2 vectors | `mapply()`, `Map()`, then `is.*()` to check type | `pmap_lgl()` (logical), `pmap_int()` (integer), `pmap_dbl()` (double), `pmap_chr()` (character), `pmap_raw()` (raw)
Side effect only | 1 vector | loops | `walk()`
Side effect only | 2 vectors | loops | `walk2()`
Side effect only | >2 vectors | loops | `pwalk()`
Data frame (`rbind` outputs) | 1 vector | `lapply()` then `rbind()` | `map_dfr()`
Data frame (`rbind` outputs) | 2 vectors | `mapply()`/`Map()` then `rbind()` | `map2_dfr()`
Data frame (`rbind` outputs) | >2 vectors | `mapply()`/`Map()` then `rbind()` | `pmap_dfr()`
Data frame (`cbind` outputs) | 1 vector | `lapply()` then `cbind()` | `map_dfc()`
Data frame (`cbind` outputs) | 2 vectors | `mapply()`/`Map()` then `cbind()` | `map2_dfc()`
Data frame (`cbind` outputs) | >2 vectors | `mapply()`/`Map()` then `cbind()` | `pmap_dfc()`
Any | Vector and its names | `l/s/vapply(X, function(x) FUN(x, names(x)))` or `mapply/Map(FUN, X, names(X))` | `imap()`, `imap_*()` (`lgl`, `dbl`, `dfr`, and etc. just like for `map()`, `map2()`, and `pmap()`)
Any | Selected elements of the vector | `l/s/vapply(X[index], FUN, ...)` | `map_if()`, `map_at()`
List | Recursively apply to list within list | `rapply()` | `map_depth()`
List | List only | `lapply()` | `lmap()`, `lmap_at()`, `lmap_if()`

### Shorthands

When an anonymous function is required in `*apply` or `map` functions, purrr offers shorthands to make the anonymous function more readable and easier to write. However, using shorthands is not recommended inside functions and packages. Here `l` denotes a list of arguments, and `f` denotes some expression involving arguments of the anonymous function.

Input | base R | purrr
----|--------|---------
1 vector | `function(x) f(x)` | `~ f(.x)`
2 vectors | `function(x, y) f(x, y)` | `~ f(.x, .y)`
More than 2 vectors | `function(x, y, z, ...) f(x, y, z, ...)` or `function(l) f(l[[1]], l[[2]], l[[3]], ...)` or `function(l) do.call(f, args = l)` | `~ f(..1, ..2, ..3, ...)`
Extract from list of vectors the `i`th element of each vector in the list | `lapply(X, function(x) tryCatch(x[["a"]], error = function(e) NA))`, `lapply(X, function(x) tryCatch(x[[3]], error = function(e) NA))` | `map(X, "a", default = NA)`, `map(X, 3, .default = NA)`

### Predicates

Here `.p`, a predicate, denotes a function that returns `TRUE` or `FALSE` indicating whether an object fulfills a criterion, such as `is.character`, and

Description | base R | purrr
------------|--------|---------
Find the value or position of the first match | `X[.p(X)][1]`, `which(.p(X))[1]` | `detect()`, `detect_index()`
Do every or some elements of a list satisfy a predicate? | For atomic vectors, `all/any(.p(X))`; for lists, `all/any(sapply(X, .p))` | `every()`, `some()`
Does a list contain an object? | `any(sapply(X, function(x) x == object))` | `has_element()`
Keep or discard elements using a predicate function | For atomic vectors, `X[.p(X)]`, `X[!.p(X)]`; for lists, `X[sapply(X, .p)]`, `X[!sapply(X, .p)]`, `X[sapply(X, function(x) length(.p(x)) > 0)]` | `keep()`, `discard()`, `compact()`
Negate a predicate function | `function(x) !.p(x)` | `negate()`

### Plucking

Extracting or modifying elements of a list

Description | base R | purrr
------------|--------|---------
Extract an element from a vector or environment, can be recursive | `[`, `[[`, or a chain of these | `pluck()`, `chuck()`
Modify an element of a vector or environment | `X[[index]] <- y` | `modify_in()`, `assign_in()`
Create an attribute getter function | `function(x) attr(x, which = a)` | `attr_getter()`

### Other vector transforms

Description | base R | purrr
------------|--------|---------
Accumulate intermediate results of a vector reduction | `Reduce(FUN, X, accumulate = TRUE)` | `accumulate()` (`FUN` takes 2 arguments), `accumulate2()` (`FUN` takes 3 arguments)
Produce all combinations of list elements | `expand.grid()` | `cross()`, `cross2()`, `cross3()`, `cross_df()`
Flatten a list of lists into a simple vector | `unlist(X, recursive = FALSE)` | `flatten()`, or with type specificity, `flatten_lgl()`, `flatten_int()`, `flatten_dbl()`, `flatten_chr()`, `flatten_raw()`, `flatten_dfr()`, `flatten_dfc()`
Recursively combine two lists | `c(X, Y)`, but more complicated to merge recursively | `list_merge()`, `list_modify()`
Reduce a list to a single value by iteratively applying a binary function | `Reduce(FUN, X, accumulate = FALSE)` | `reduce()`, `reduce2()`

### Adverbs

Adverbs modify the action of a function; taking a function as input and returning a function with modified action as output.

Description | base R | purrr
------------|--------|-----
Compose functions | `function(x) f1(f2(f3(x)))` |`compose()`
Lift the domain of a function | `function(...) FUN(list(...))`, `function(l) do.call(FUN, l)` | `lift()`, `lift_dl()`, `lift_dv()`, `lift_vl()`, `lift_vd()`, `lift_ld()`, `lift_lv()`
Partial apply a function, filling in some arguments | `function(x) FUN(x, y = foo)` | `partial()`
Handling errors, warnings, and messages | `try()`, `tryCatch()`, `withCallingHandlers()` | `safely()`, `quietly()`, `possibly()`, `auto_browse()`

### Misc

Description | base R | purrr
------------|--------|---------
Default value for `NULL` | `function(x, y) if (is.null(x)) return(y) else return(x)` | `%||%`
Coerce array to list | For matrices, `as.data.frame`, perhaps after transposing, as data frames are lists | `array_tree()`, `array_branch()`
Coerce a list to a vector | `unlist()` | `as_vector()`, `simplify()` (simplify if possible), `simplify_all()` (recursively simplify); the purrr version can have type specificity.
Generate random sample from a Bernoulli distribution | `function(n, p) as.logical(rbinom(n = n, size = 1, prob = p))` | `rbernoulli()`
Generate random sample from a discrete uniform distribution | `function(n, a, b) sample(a:b, size = n)` | `rdunif()`
Set names of a vector | `setNames()` | `set_names()`, but has more features and stricter than `setNames()`

## Examples

### Applying function to one vector

Suppose we would like to generate a list of samples of 5 from normal distributions with the same standard deviation but different means. We want a list of numerical vectors. Then we would like to find the median of each sample, which should be a numeric vector. Then we will plot a histogram for each sample.

Take the samples, base R approach
```{r}
set.seed(2020)
means <- 1:4
(samples <- lapply(means, rnorm, n = 5, sd = 1))
```

Take samples, purrr approach. Here `map()` is just like `lapply()`, always returning a list, no simplification.
```{r}
set.seed(2020)
(samples <- map(means, rnorm, n = 5, sd = 1))
```

Compute the medians, base R approach

```{r}
# type stable
(medians <- vapply(samples, median, FUN.VALUE = numeric(1L)))

# not type stable
(medians <- sapply(samples, median))
```

Compute the medians, purrr approach

```{r}
(medians <- map_dbl(samples, median))
```

Here `map_dbl()` ensures that the the output is double, throwing an error if the output is of other types.

How about when we simply want the side effect when applying a function on a vector, such as a plot or a file output, but not the returned values? 

Side effect, base R approach:
```{r}
par(mfrow = c(2,2))
for (s in samples) {
  hist(s, xlab = "value", main = "")
}
```

Side effect, purrr approach:

```{r}
par(mfrow = c(2,2))
walk(samples, ~ hist(.x, xlab = "value", main = ""))
```


### Applying function to more than one vector

Suppose we would like to take samples from normal distributions with different means and standard deviations. We need to apply `rnorm()` to two vectors, one for means, the other for standard deviations.

```{r}
means <- 1:4
sds <- 1:4
```

Get samples, base R approach:
```{r}
set.seed(2020)
(samples <- mapply(rnorm, mean = means, sd = sds, MoreArgs = list(n = 5), SIMPLIFY = FALSE))
```

Get samples, purrr approach
```{r}
set.seed(2020)
(samples <- map2(means, sds, rnorm, n = 5))
```

Calculating the median would be the same as the last section.
```{r}
(medians <- map_dbl(samples, median))
```

### Data frame outputs

Suppose the output is a list data frames with the same column names; it makes sense to combine these data frames into a single data frame.

Base R approach:
```{r}
set.seed(2020)
dfs <- lapply(means, function(m) {
  tibble(
    x = rnorm(3, mean = m, sd = 1),
    y = rnorm(3, mean = m + 1, sd = 2),
    z = sample(LETTERS, 1)
  )
})

df <- do.call(rbind, dfs)
df
```

Then the purrr approach:
```{r}
set.seed(2020)
df <- map_dfr(means, ~ tibble(
  x = rnorm(3, mean = .x, sd = 1),
  y = rnorm(3, mean = .x + 1, sd = 2),
  z = sample(LETTERS, 1)
))
df
```

