---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

purrr is a toolkit that helps you write cleaner, more maintainable R code by introducing functional programming concepts. But what is functional programming? At its core, it's an approach to programming that emphasizes using functions to transform data, similar to how you might use a series of tools to process raw materials into a final product.

Instead of writing loops and modifying data step by step, functional programming encourages you to think about your data transformations as a series of function applications. This might sound abstract, but we'll see how it makes your code clearer and less prone to errors.

## Installation and Setup

```{r setup}
# Install purrr if you haven't already
# install.packages("purrr")

# Load the package
library(purrr)
```

## Understanding Lists in R

Before diving into purrr, let's understand lists, as they're fundamental to how purrr works:

```{r}
# A list can contain different types of data
my_list <- list(
  data_frame = data.frame(x = 1:3, y = 4:6),
  model = lm(mpg ~ wt, mtcars),
  custom_function = function(x) x^2
)

# Access list elements
my_list$data_frame  # Using $
my_list[[1]]       # Using [[]] to get the first element
my_list[1]         # Using [] returns a list with one element
```

## Basic Map Functions: A Better Way to Loop

### map(): The Foundation

Instead of writing loops, map() provides a cleaner way to apply functions to each element of a list:

```{r}
# Process multiple texts with complex operations
documents <- list(
  "The quick brown fox",
  "jumps over the lazy dog",
  "and then takes a nap"
)

# Complex text processing
processed_docs <- map(.x = documents, .f = function(doc) {
  # Remove punctuation
  clean <- gsub("[[:punct:]]", "", doc)
  # Convert to lowercase
  clean <- tolower(clean)
  # Get unique words
  words <- unique(strsplit(clean, "\\s+")[[1]])
  # Create frequency table
  table(strsplit(clean, "\\s+")[[1]])
})

# Process multiple data frames with custom operations
data_frames <- list(
  df1 = data.frame(x = c(1, 2, NA), y = c("a", NA, "c")),
  df2 = data.frame(x = c(4, NA, 6), y = c(NA, "e", "f"))
)

summaries <- map(.x = data_frames, .f = function(df) {
  # Calculate custom summary statistics
  numeric_cols <- sapply(df, is.numeric)
  list(
    missing_by_col = sapply(df, function(col) sum(is.na(col))),
    complete_cases = sum(complete.cases(df)),
    numeric_summary = if(any(numeric_cols)) {
      sapply(df[numeric_cols], function(col) {
        c(mean = mean(col, na.rm = TRUE),
          sd = sd(col, na.rm = TRUE))
      })
    }
  )
})
```

### The Progress Bar

map() has options for enabling the progress bar:

```{r}
# Rerun the previous example with the progress bar enabled
summaries <- map(.x = data_frames, .f = function(df) {
  # Calculate custom summary statistics
  numeric_cols <- sapply(df, is.numeric)
  list(
    missing_by_col = sapply(df, function(col) sum(is.na(col))),
    complete_cases = sum(complete.cases(df)),
    numeric_summary = if(any(numeric_cols)) {
      sapply(df[numeric_cols], function(col) {
        c(mean = mean(col, na.rm = TRUE),
          sd = sd(col, na.rm = TRUE))
      })
    }
  )
}, .progress = TRUE)

```

### Type-specific Map Functions

Here are examples where type-specific mapping is useful for non-vectorized operations:

```{r}
# Calculate complexity metrics for each document
complexity_scores <- map(.x = documents, .f = function(doc) {
  words <- strsplit(doc, "\\s+")[[1]]
  unique_ratio <- length(unique(words)) / length(words)
  avg_word_length <- mean(nchar(words))
  unique_ratio * avg_word_length
})

# Generate quality reports for data frames
quality_reports <- map(.x = data_frames, .f = function(df) {
  issues <- character(0)
  
  # Check for missing values
  if(any(is.na(df))) {
    issues <- c(issues, 
      sprintf("Missing values: %d", sum(is.na(df))))
  }
  
  # Check for duplicates
  if(any(duplicated(df))) {
    issues <- c(issues,
      sprintf("Duplicate rows: %d", sum(duplicated(df))))
  }
  
  # Combine all issues
  if(length(issues) > 0) {
    paste("Issues found:", paste(issues, collapse = "; "))
  } else {
    "No issues found"
  }
})
```

## Function Composition with the Base Pipe

Here's how to chain operations that can't be vectorized:

```{r}
# Process multiple datasets with different steps
models <- list(
  lm(mpg ~ wt, mtcars),
  lm(mpg ~ wt + cyl, mtcars),
  lm(mpg ~ ., mtcars)
)

model_analysis <- models |>
  map(.f = function(model) {
    # Extract model diagnostics
    residuals <- residuals(model)
    fitted <- fitted(model)
    
    list(
      residuals = residuals,
      fitted = fitted,
      stats = summary(model)
    )
  }) |>
  map(.f = function(analysis) {
    # Perform residual diagnostics
    with(analysis, {
      list(
        normality = shapiro.test(residuals)$p.value,
        r_squared = stats$r.squared
      )
    })
  }) |>
  map(.f = function(diagnostics) {
    # Create formatted summary
    sprintf(
      "Normality p=%.3f, R²=%.3f",
      diagnostics$normality,
      diagnostics$r_squared
    )
  })
```

## Working with Multiple Inputs

### map2(): Parallel Processing

Here are examples where you need to process pairs of inputs with non-vectorized operations:

```{r}
# Compare pairs of models
model_pairs <- map2(.x = models[-1], .y = models[-length(models)], .f = function(current, previous) {
  # Compare model fits
  anova_result <- anova(previous, current)
  
  # Extract comparison metrics
  list(
    df_diff = diff(anova_result$Df),
    deviance_diff = diff(anova_result$Deviance),
    p_value = anova_result$`Pr(>F)`[2]
  )
})

# Process pairs of data frames with custom comparisons
df_comparisons <- map2(.x = data_frames[-1], .y = data_frames[-length(data_frames)], 
  .f = function(current, previous) {
    # Compare structure and content
    list(
      added_cols = setdiff(names(current), names(previous)),
      removed_cols = setdiff(names(previous), names(current)),
      common_cols = intersect(names(current), names(previous))
    )
  }
)
```

### pmap(): Multiple Inputs

For operations requiring multiple parameters that can't be vectorized:

```{r}
# Configure and run multiple analyses
analysis_configs <- list(
  data = list(mtcars, PlantGrowth, InsectSprays),
  dependent = list("mpg", "weight", "count"),
  predictors = list(
    c("wt", "cyl"),
    c("group"),
    c("spray")
  ),
  test_type = list("lm", "aov", "glm")
)

analysis_results <- pmap(.l = analysis_configs, .f = function(data, dependent, predictors, test_type) {
  # Construct formula
  formula_str <- paste(dependent, "~", paste(predictors, collapse = " + "))
  formula <- as.formula(formula_str)
  
  # Fit model based on test type
  model <- switch(test_type,
    "lm" = lm(formula, data = data),
    "aov" = aov(formula, data = data),
    "glm" = glm(formula, data = data)
  )
  
  # Return analysis results
  list(
    model_type = test_type,
    formula = formula_str,
    summary = summary(model),
    diagnostics = list(
      residuals = residuals(model),
      fitted = fitted(model)
    )
  )
})
```

## Error Handling

When processing real data, things often go wrong. purrr helps you handle errors gracefully:

```{r}
# A complex analysis function that might fail
analyze_dataset <- function(df) {
  if (!all(c("x", "y") %in% names(df))) {
    stop("Missing required columns x and y")
  }
  
  if (!all(sapply(df[c("x", "y")], is.numeric))) {
    stop("x and y must be numeric")
  }
  
  # Perform analysis
  model <- lm(y ~ x, data = df)
  residuals <- residuals(model)
  
  list(
    coefficients = coef(model),
    r_squared = summary(model)$r.squared,
    normality = shapiro.test(residuals)$p.value
  )
}

# List of datasets to analyze
datasets <- list(
  data.frame(x = 1:5, y = 1:5),                    # Valid
  data.frame(x = "a", y = 1:5),                    # Invalid types
  data.frame(z = 1:5),                             # Missing columns
  data.frame(x = c(1, 1, 1), y = c(2, 2, 2))      # Perfect correlation
)

# Using safely()
safe_analysis <- safely(analyze_dataset)
results <- map(.x = datasets, .f = safe_analysis)
# Each element contains $result and $error

# Using possibly()
safe_analysis2 <- possibly(analyze_dataset, otherwise = list(
  coefficients = NA,
  r_squared = NA,
  normality = NA
))
results2 <- map(.x = datasets, .f = safe_analysis2)

# Using quietly()
# quiet_analysis <- quietly(analyze_dataset)
# results3 <- map(.x = datasets, .f = quiet_analysis)
# Each element contains $result, $output, $warnings, and $messages
```

## Practical Examples

### Example 1: Advanced Data Frame Operations

```{r}
# Process multiple data frames with complex operations
data_frames <- list(
  sales = data.frame(
    date = as.Date("2024-01-01") + 0:9,
    amount = runif(10, 100, 1000),
    category = sample(c("A", "B", "C"), 10, replace = TRUE)
  ),
  inventory = data.frame(
    date = as.Date("2024-01-01") + 0:9,
    item = sample(letters[1:5], 10, replace = TRUE),
    quantity = rpois(10, 20)
  )
)

processed_data <- data_frames |>
  map(.f = function(df) {
    # Add custom datetime features
    df$year <- format(df$date, "%Y")
    df$month <- format(df$date, "%m")
    df$day <- format(df$date, "%d")
    df$weekday <- format(df$date, "%A")
    
    # Calculate statistics if numeric columns exist
    numeric_cols <- sapply(df, is.numeric)
    if (any(numeric_cols)) {
      numeric_data <- df[numeric_cols]
      df$mean <- apply(numeric_data, 
                       MARGIN = 2,
                       FUN = mean)
    }
    
    df
  }) |>
  map(.f = function(df) {
    # Calculate summary statistics by group if categorical columns exist
    categorical_cols <- sapply(df, is.factor) | sapply(df, is.character)
    if (any(categorical_cols)) {
      group_col <- names(df)[categorical_cols][1]
      numeric_cols <- sapply(df, is.numeric)
      
      summaries <- aggregate(df[numeric_cols], 
                           by = list(group = df[[group_col]]), 
                           FUN = function(x) {
                             c(mean = mean(x, na.rm = TRUE),
                               sd = sd(x, na.rm = TRUE))
                           })
    }
    
    list(
      processed_data = df,
      group_summaries = if (exists("summaries")) summaries else NULL
    )
  })
```

### Example 2: Model Evaluation Pipeline

```{r}
# Create and evaluate multiple models
model_specs <- list(
  list(formula = mpg ~ wt,           name = "simple"),
  list(formula = mpg ~ wt + cyl,     name = "medium"),
  list(formula = mpg ~ .,            name = "complex")
)

model_results <- model_specs |>
  map(.f = function(spec) {
    # Fit model
    model <- lm(spec$formula, data = mtcars)
    
    # Perform cross-validation
    cv_results <- lapply(1:5, \(fold) {
      # Create train/test split
      test_idx <- seq_len(nrow(mtcars)) %% 5 == (fold - 1)
      train <- mtcars[!test_idx, ]
      test <- mtcars[test_idx, ]
      
      # Fit on train, predict on test
      fold_model <- lm(spec$formula, data = train)
      predictions <- predict(fold_model, newdata = test)
      
      # Calculate fold metrics
      mse <- mean((test$mpg - predictions)^2)
      mae <- mean(abs(test$mpg - predictions))
      
      list(mse = mse, mae = mae)
    })
    
    # Combine results
    list(
      name = spec$name,
      formula = deparse(spec$formula),
      model = model,
      cv_metrics = list(
        mse = mean(sapply(cv_results, function(x) x$mse)),
        mae = mean(sapply(cv_results, function(x) x$mae))
      ),
      diagnostics = list(
        r_squared = summary(model)$r.squared,
        aic = AIC(model),
        residual_normality = shapiro.test(residuals(model))$p.value
      )
    )
  })
```

## Best Practices

1. Use map functions when:

- Processing complex objects (like models or data frames)
- Performing operations that can't be vectorized
- Handling errors and edge cases
- Working with nested data structures

2. Avoid map functions when:

- Working with simple arithmetic operations
- Using functions that are already vectorized
- Processing single values or simple vectors

3. Error Handling:

- Use safely() when you need to know what went wrong
- Use possibly() when you just want a default value on error
- Use quietly() when you need to capture messages and warnings

4. Code Organization:

- Break complex operations into smaller functions
- Use meaningful names for intermediate steps
- Comment your code to explain the "why", not just the "what"

## Common Mistakes to Avoid

1. Using map for vectorized operations:

```{r}
# Don't do this
map(.x = 1:10, function(x) x * 2)  # Use regular vectorization instead: 1:10 * 2

# Do this
data_frames <- list(
  sales = data.frame(
    date = as.Date("2024-01-01") + 0:9,
    y = runif(10, 100, 1000),
    x = sample(c("A", "B", "C"), 10, replace = TRUE)
  ),
  inventory = data.frame(
    date = as.Date("2024-01-01") + 0:9,
    x = sample(letters[1:5], 10, replace = TRUE),
    y = rpois(10, 20)
  )
)

map(.x = data_frames, .f = function(df) {
  # Complex non-vectorized operations
  model <- lm(y ~ x, data = df)
  summary(model)$r.squared
})
```

2. Neglecting error handling:

```{r}
# Don't do this
# map(.x = datasets, .f = analyze_dataset)  # Might fail on invalid data

# Do this
map(.x = datasets, .f = safely(analyze_dataset))  # Handles errors gracefully
```

3. Making functions too complex:

```{r}
# Don't do this
# map(data)
```

