---
title: "Iterating over rows using pmap"
author:
- Safouane Chergui
- Jenny Bryan
---

### Goals

The main goal of this vignette is to show how one can interate over rows using functions from purrr package. [Jenny Bryan's repository](https://github.com/jennybc/row-oriented-workflows) is the main inspiration of this vignette. The reader is encouraged to check Jenny Bryan's github repository after reading this vignette as it has more examples and goes into more details.

In this vignette, we are going to:

- Arrange inputs as rows in a data frame

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(purrr)
```

### Generate numbers with different parameters  via `runif()`

Use `pmap()` in order to iterate over several arguments rowwise. This function comes in handy when one needs to generate numbers from a distribution while passing different argument each time the function gets called.

The following chunk of code r `runif(n, min = 0, max = 1)` generates n numbers from the uniform distribution. The same operation can be done for several triples. 
The first step is to store triples in a dataframe and then iterate over rows.

```{r}
# Notice how df's variable names are the same as runif's argument names.
df <- tribble(
  ~ n, ~ min, ~ max,
   1L,     0,     1,
   2L,    10,   100,
   3L,   100,  1000
)
df
```

Let us first set the seed and practice on single rows. Setting the seed will make the result reproducible and practicing on single rows will allow us to check if the results we get with `pmap()` are what we truly expect to get.

```{r}
set.seed(123)
(x <- df[1, ])

runif(n = x$n, min = x$min, max = x$max)

x <- df[2, ]
runif(n = x$n, min = x$min, max = x$max)

x <- df[3, ]
runif(n = x$n, min = x$min, max = x$max)

```

Now, let us try to do the same thing with `pmap()`:

```{r}
set.seed(123)
pmap(df, runif)
```

### Finessing variable and argument names

It is not mandatory for variable names to be the same as the argument names of the function. This can be arranged on the fly.In the example below, we do exactly this.

```{r}
# Create a tibble with three columns whose names are different from runif argument names
foofy <- tibble(
  num_generated = 1:3,            ## was: n
  minimum = c(0, 10, 100),   ## was: min
  maximum = c(1, 100, 1000) ## was: max
)
```

Let us rename variables on the fly using the pipe operator.

```{r}
set.seed(123)
foofy %>%
  rename(n = num_generated, min = minimum, max = maximum) %>%
  pmap(runif)
```

### Extra variables in the data frame

One situation that comes up so often is having a dataframe that includes columns that shouldn't be passed to the function `.f()`. Let us create a tibble to illustrate this.

```{r}
df_oops <- tibble(
  n = 1:3,
  min = c(0, 10, 100),
  max = c(1, 100, 1000),
  oops = c("please", "ignore", "me")
)
df_oops
```

Passing the dataframe as it is is not going to work !

```{r error = T}
set.seed(123)
pmap(df_oops, runif)
```

Selecting the specific columns that should be passed to `.f()` or ommitting the columns that shouldn't be passed is a good way of doing. This can be done using `dplyr::select()`.

```{r}
set.seed(123)
df_oops %>%
  select(n, min, max) %>% ## if it's easier to say what to keep
  pmap(runif)
```

One can also omit columns if it turns out to be easier.

```{r eval=F}
set.seed(123)
df_oops %>%
  select(-oops) %>%       ## if it's easier to say what to omit
  pmap(runif)
```

A custom wrapper that uses `...` to ignore additional arguments can also get the job done.

```{r}
my_runif <- function(n, min, max, ...) runif(n, min, max)

set.seed(123)
pmap(df_oops, my_runif)
```

### Add the generated data to the data frame as a list-column

The generated data can be embedded inside the dataframe using `mutate()`.

```{r}
set.seed(123)
# The `.` comes from magrittr. It's only going to work if used in a pipe.
# Using `.` when df is grouped might give surprising results.
(df_aug <- df %>%
    mutate(data = pmap(., runif)))
```

This can still be done if the name of the columns aren't the same as the name of the function arguments. `list()` should be used instead of the placeholder `.`. Variables can then be renamed inside the list.

```{r}
foofy <- tibble(
  alpha = 1:3,            ## was: n
  beta = c(0, 10, 100),   ## was: min
  gamma = c(1, 100, 1000) ## was: max
)

set.seed(123)
foofy %>%
  mutate(data = pmap(list(n = alpha, min = beta, max = gamma), runif))
```

`list()`can also be used to select only the variables that are needed or to exclude only some of the variables.

```{r}
df_oops <- tibble(
  n = 1:3,
  min = c(0, 10, 100),
  max = c(1, 100, 1000),
  oops = c("please", "ignore", "me")
)

set.seed(123)
df_oops %>%
  mutate(data = pmap(list(n, min, max), runif))

# Excluding "oops" column
df_oops %>%
  mutate(data = pmap(select(., -oops), runif))
```


